Optimal updating factor in adaptive flat-distribution simulations

### 2016-04-29 doc/invt.tex
add data for Table I
tikz version of Figure 1.

\documentclass[reprint, floatfix]{revtex4-1}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{upgreek}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{tikz}
\usepackage{hyperref}

@@

To simplify the ensuing discussion,
we shall further impose the
detailed balance condition:
%
\begin{equation}
  p_i \, w_{ij} = p_j \, w_{ji}.
  \label{eq:w_detailedbalance}
\end{equation}
%
It follows that
\begin{equation}
  \sum_{i = 1}^n p_i \, w_{ij}
  =
  \sum_{i = 1}^n p_j \, w_{ji}
  = p_j,
  \label{eq:w_balance}
\end{equation}
%
i.e., $\mathbf p$ is a left eigenvector of
$\mathbf w$ with eigenvalue $1$.

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.9\linewidth]{fig/massq.pdf}
  }
  \caption{
    \label{fig:massq}
    Geometric characterization of the optimal schedule,
    $\alpha(t)$,
    via the normalized mass distribution, $m(Q)$.
    %
    \note{The figure was produced by \texttt{doc/fig/massq.gp}.
    }%
  }
\end{center}
\end{figure}

@@

Let us now consider simulations in which
the system is initially equilibrated at the same
constant updating magnitude $\alpha$.
%
For these simulations,
we can adjust $q(T)$
to minimize the total error.
%
From Eqs. \eqref{eq:error_res1}
and \eqref{eq:error_asym2},
we find that
the $q(T)$ that minimizes
the total error satisfies
\begin{equation}
  m\bigl( q(T) \bigr)
  =
  \frac{1} { T \, (\alpha_0 / 2) }
  .
\label{eq:opt_qT}
\end{equation}
%
\note{Let $Q_c \equiv q(T)$,
$$
\begin{aligned}
  \frac{
    \partial E_R
  }
  {
    \partial q(T)
  }
  &=
  -\alpha_0 \, M^2(Q_c)
  ,
  \\
  \frac{
    \partial E_A
  }
  {
    \partial q(T)
  }
  &=
  \frac 2 T \,
  M(Q_c) \,
  \int_0^{ Q_c } M(Q) \, dQ
  .
\end{aligned}
$$
Then solve $\partial (E_R + E_A) / \partial q(T) = 0$.
}
%
Then, by Eqs. \eqref{eq:mQ_invTa} and \eqref{eq:intmQ_tT},
we have
\begin{equation}
  \alpha( t = 0 )
  =
  \frac{ \alpha_0 }
       { 2 }
  ,
\label{eq:half_alpha0}
\end{equation}
%
i.e., the optimal initial updating magnitude
is always half of the equilibration value.
%
This factor $1/2$
happens to coincide with the
reduction factor of updating magnitude
at stage transitions
in the original WL algorithm\cite{
wang2001, wang2001pre}.

@@

Many commonly-used updating schemes
can be characterized by a rigid window
centered on the current bin.
%
We call these schemes homogeneous,
or translationally invariant.
%
An example is the Gaussian updating scheme
used in metadynamics.
%
Homogeneous updating schemes
share the same set of eigenvectors
but not the corresponding eigenvalues.
%
They have significantly fewer
parameters than the general case,
and the multiplication of the updating matrix,
$\mathbf w$,
can be reduced to a convolution with the window,
or the updating kernel.
%
Thus,
there is a one-to-one correspondence between
the eigenvalues of $\mathbf w$
and the updating kernel.

@@

While translationally-invariant
updating schemes
naturally suit periodic variables\cite{
dama2014},
they can also be extended to non-periodic variables
by imposing the reflective boundary condition\cite{
bussi2006}.
%
For simplicity, we shall assume that the target
distribution is flat, or $p_i = 1/n$, below.

It is natural to introduce homogeneous updating schemes
for a periodic variable\cite{dama2014}.
%
The updating matrix, $\mathbf w$,
assume the following form
%
\begin{equation}
  w_{ij}
  =
  \mu_{i-j}
  +
  \mu_{i-j+n}
  +
  \mu_{i-j-n}
  ,
\notag
%\label{eq:w_band_pbc}
\end{equation}
%
where
$\mu_{-b}, \dots, \mu_b$ ($b \le n/2$)
are a set of numbers satisfying
%
\begin{equation}
  \mu_{-b} + \cdots + \mu_b = 1
  ,
\label{eq:mu_normalization}
\end{equation}
%
and $\mu_l = 0$ for $|l| > b$.
%
Particularly,
we shall only consider symmetric kernels
that satisfy
%
\begin{equation}
  \mu_i = \mu_{-i}
  .
\label{eq:mu_symm}
\end{equation}
%
If $b = n/2$, then $\mu_{-b}$ is removed
from the sum in Eq. \eqref{eq:mu_normalization}
to avoid double counting.
%
These numbers constitutes an \emph{updating kernel}.

@@

%Since the error depends on
%the underlying sampling process,
We tested two sampling processes.
%a global one and a local one.
%
Both are based on
Metropolis MC algorithm\cite{
  metropolis1953, newman, frenkel,
  landau_binder}:
%
in each step, a new bin index $j$ is proposed
and then accepted with probability
%
$
A(i \to j) = \min\{ 1, \exp(V_i - V_j) \}.
$
However,
in the first or global sampling process,
we choose the destination $j$
uniformly out of the $n$ bins.
%
In the second or local sampling process,
$j$ is either $i - 1$ or $i + 1$
(adjusted by the boundary condition)
with equal probability $1/2$.
%
Asymptotically,
the above global and local sampling processes
emulate the perfect and one-step processes
discussed in Appendix \ref{sec:Gamma},
respectively,
and we shall use
Eqs. \eqref{eq:Gamma_perfect}
and \eqref{eq:Gamma_onestep}
for the respective $\Gamma_k$'s.
%

@@

\begin{table}[h]
  \caption{\label{tab:error_Gaussian}
    Error of the Gaussian updating scheme of $\sigma = 10$.
    $t_0$ is given by Eq. \eqref{eq:t0_sinc}.
    \note{ Use the simulation data.  }%
  }
  \setlength{\tabcolsep}{5pt}
  \begin{tabular} { l | p{2cm} p{2.5cm} p{1.8cm} }
    \hline
      Schedule
    &
      Inverse-time, Eq. \eqref{eq:alpha_invt1}
    &
      Modified inverse-time,
      Eq.~\eqref{eq:alpha_invtlambda}
      with optimized $\lambda$
    &
      Optimal,
      Eq. \eqref{eq:q_opt}
    \\
    \hline
    Global
    &
    $4.2 \times 10^{-6}$
    &
    $7.1 \times 10^{-7}$
    &
    $2.7 \times 10^{-7}$
    \\
    \hline
    Local
    &
    $3.7 \times 10^{-4}$
    &
    $1.7 \times 10^{-4}$
    &
    $9.1 \times 10^{-5}$
    \\
    \hline
  \end{tabular}
\end{table}

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=1.0\linewidth]{fig/xerr.pdf}
  }
  \caption{
    \label{fig:xerr}
    Error components for the Gaussian updating scheme
    with $\sigma = 10$
    from the optimal schedule
    and the inverse-time schedule, Eq. \eqref{eq:alpha_invt1}.
    %
    The latter schedule failed to damp out the errors
    from the long-wavelength (local) eigenmodes
    with $K \ge 5$.
    %
    For longer simulations,
    the error components from the optimal schedule
    would roughly maintain the same shape,
    but ``intersect'' with the initial error curve
    at a lower level (not shown).
    %
    \note{This figure was produced by \texttt{doc/fig/xerr.gp}.
      The data were saved under the directory \texttt{data/xerr}.
    }%
  }
\end{center}
\end{figure}

@@

To show Eq. \eqref{eq:varY},
we first use Eq. \eqref{eq:v_def}
to rewrite Eq. \eqref{eq:x_def}
as
$
  x_i
  =
  (V_i - \bar V) + s_i,
$
where
$
  s_i
  \equiv
  \ln \frac{ p_i } { p^*_i }
  -
  \sum_{ j = 1 }^n
    p_j \, \ln \frac{ p_j } { p^*_j }.
$
Thus, by Eqs. \eqref{eq:y_def} and \eqref{eq:Y_def},
we have
$$
y_k = Y_k + \sum_{i = 1}^n \phi_{k i} \, s_i.
$$
Since the second term on the right-hand side
is a constant, and the average of $y_k$ is zero,
we have
$$
\operatorname{var} Y_k
=
\operatorname{var} y_k
=
\langle y_k^2 \rangle.
$$
Finally, by using Eq. \eqref{eq:y2_eql},
we reach Eq. \eqref{eq:varY}.

@@

As an example, we used the method to compute
the integrals, $\Gamma_k$'s, from an MD simulation,
and compared them to those from
Eq. \eqref{eq:Gamma_onestep}.
%the one-step sampling process.
%
As shown in Fig. \ref{fig:gamcmp},
the larger values of $\Gamma_k$'s
with smaller $k$'s
from the two sampling processes
were almost proportional.
%
This suggests that
the optimal schedule
from simulations using local sampling processes
would be similar in general properties.

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/gamcmp.pdf}
  }
  \caption{
    \label{fig:gamcmp}
    Integrals of the autocorrelation functions
    from an MD simulation versus those
    from the one-step process.
    %
    The MD simulation was ran for $10^{10}$ steps with
    the step size $\Delta t_\mathrm{MD} = 0.005$.
    %
    The bin size $\Delta z = 2\, \pi /n$ with $n = 100$,
    and the temperature was $1.0$.
    %
    The velocity rescaling thermostat\cite{bussi2007} was
    applied every MD step with viscosity $20$.
    %
    The reflective boundary condition was used.
    %
    \note{The figure was produced by \texttt{doc/fig/gamcmp.gp}.
      The data were saved under \texttt{data/gamcmp}
      (see the \texttt{Makefile} there).
    }%
  }
\end{center}
\end{figure}

>>>>

\documentclass[reprint, floatfix]{revtex4-1}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{upgreek}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{tikz}
\usetikzlibrary{calc}
\usepackage{hyperref}

@@

To simplify the ensuing discussion,
we shall further impose the
detailed balance condition:
%
\begin{equation}
  \rho_i \, w_{ij} = \rho_j \, w_{ji}.
  \label{eq:w_detailedbalance}
\end{equation}
%
It follows that
\begin{equation}
  \sum_{i = 1}^n p_i \, w_{ij}
  =
  \sum_{i = 1}^n p_j \, w_{ji}
  = p_j,
  \label{eq:w_balance}
\end{equation}
%
i.e., $\mathbf p$ is a left eigenvector of
$\mathbf w$ with eigenvalue $1$.

@@

\begin{figure}[h]
\begin{center}
  \begin{tikzpicture}[very thick, scale=2.5, every node/.style={scale=1.5}]
    % distribution
    \draw[domain=0:2.5, variable=\x, dashed, gray, ultra thick] 
      plot({\x}, {1.5*exp(-0.5*\x*\x)});

    % shade
    \fill [gray!30!white, domain=0.7:2.2, variable=\x]
      (0.7, 0) --
      plot ({\x}, {1.5*exp(-0.5*\x*\x)}) 
      -- (2.2, 0) -- cycle;

    % thick curve
    \draw[domain=0:2.2, variable=\x, ultra thick]
      plot({\x}, {1.5*exp(-0.5*\x*\x)});

    % label for the shaded area
    \node[] at (1.1, 0.5) {$\dfrac{t}{T}$};

    \draw[<->] (0.6, 0)  --
               node[left, inner sep=1mm]
               {$\dfrac{1}{T \, \alpha}$}
               (0.6, 1.174);
    % label for the ordinate
    %\node[fill=white] at (0.4, 0.7) {$\dfrac{1}{Ta}$};

    \draw[] (0.5, 1.174) -- (0.7, 1.1740);
    \draw[->] (0, 0) -- node[below, inner sep=5mm] {$Q$} (2.8, 0);
    \draw[->] (0, 0) -- node[above, rotate=90] {$M(Q)$} (0, 1.8);

    % xtics
    \draw[] (0.0, 0.1) -- (0.0, 0.0) node[below] {$0$};
    \draw[] (2.2, 0.1) -- (2.2, 0.0) node[below] {$q(T)$};
  \end{tikzpicture}
  %\makebox[\linewidth][c]{
  %  \includegraphics[angle=0, width=0.9\linewidth]{fig/massq.pdf}
  %}
  %\caption{
  %  \label{fig:massq}
  %  Geometric characterization of the optimal schedule,
  %  $\alpha(t)$,
  %  via the normalized mass distribution, $m(Q)$.
  %  %
  %  \note{The figure was produced by \texttt{doc/fig/massq.gp}.
  %  }%
  %}
\end{center}
\end{figure}

@@

Let us now consider simulations in which
the system is initially equilibrated at the same
constant updating magnitude $\alpha_0$.
%
For these simulations,
we can adjust $q(T)$
to minimize the total error.
%
From Eqs. \eqref{eq:error_res1}
and \eqref{eq:error_asym2},
we find that
the $q(T)$ that minimizes
the total error satisfies
\begin{equation}
  m\bigl( q(T) \bigr)
  =
  \frac{1} { T \, (\alpha_0 / 2) }
  .
\label{eq:opt_qT}
\end{equation}
%
\note{Let $Q_c \equiv q(T)$,
$$
\begin{aligned}
  \frac{
    \partial E_R
  }
  {
    \partial q(T)
  }
  &=
  -\alpha_0 \, M^2(Q_c)
  ,
  \\
  \frac{
    \partial E_A
  }
  {
    \partial q(T)
  }
  &=
  \frac 2 T \,
  M(Q_c) \,
  \int_0^{ Q_c } M(Q) \, dQ
  .
\end{aligned}
$$
Then solve $\partial (E_R + E_A) / \partial q(T) = 0$.
}%
By Eqs. \eqref{eq:mQ_invTa} and \eqref{eq:intmQ_tT},
we can rewrite the condition as
\begin{equation}
  \alpha( t = 0 )
  =
  \frac{ \alpha_0 }
       { 2 }
  ,
\label{eq:half_alpha0}
\end{equation}
%
i.e., the optimal initial updating magnitude
is always half of the equilibration value.
%
This factor $1/2$
happens to coincide with the
reduction factor of updating magnitude
at stage transitions
in the original WL algorithm\cite{
wang2001, wang2001pre}.

@@

Many commonly-used updating schemes
can be characterized by a rigid window
centered on the current bin.
%
We call these schemes homogeneous,
or translationally invariant.
%
Naturally, $\rho_i = 1/n$ in this case.
%
An example is the Gaussian updating scheme
used in metadynamics.
%
Homogeneous updating schemes
share the same set of eigenvectors
but not the corresponding eigenvalues.
%
They have significantly fewer
parameters than the general case,
and the multiplication of the updating matrix,
$\mathbf w$,
can be reduced to a convolution with the window,
or the updating kernel.
%
Thus,
there is a one-to-one correspondence between
the eigenvalues of $\mathbf w$
and the updating kernel.

@@

%While translationally-invariant
%updating schemes
%naturally suit periodic variables\cite{
%dama2014},
%they can also be extended to non-periodic variables
%by imposing the reflective boundary condition\cite{
%bussi2006}.
%%
%For simplicity, we shall assume that the target
%distribution is flat, or $p_i = 1/n$, below.

It is natural to first introduce homogeneous updating schemes
for a periodic variable\cite{dama2014}.
%
The updating matrix, $\mathbf w$,
assume the following form
%
\begin{equation}
  w_{ij}
  =
  \mu_{i-j}
  +
  \mu_{i-j+n}
  +
  \mu_{i-j-n}
  ,
\notag
%\label{eq:w_band_pbc}
\end{equation}
%
where
$\mu_{-b}, \dots, \mu_b$ ($b \le n/2$)
are a set of numbers satisfying
%
\begin{equation}
  \mu_{-b} + \cdots + \mu_b = 1
  ,
\label{eq:mu_normalization}
\end{equation}
%
and $\mu_l = 0$ for $|l| > b$.
%
Particularly,
we shall only consider symmetric kernels
that satisfy
%
\begin{equation}
  \mu_i = \mu_{-i}
  .
\label{eq:mu_symm}
\end{equation}
%
If $b = n/2$, then $\mu_{-b}$ is removed
from the sum in Eq. \eqref{eq:mu_normalization}
to avoid double counting.
%
These numbers constitutes an \emph{updating kernel}.

@@

%Since the error depends on
%the underlying sampling process,
We tested two sampling processes.
%a global one and a local one.
%
Both are based on
Metropolis MC algorithm\cite{
  metropolis1953, newman, frenkel,
  landau_binder}:
%
in each step, a new bin index $j$ is proposed
and then accepted with probability
%
$
A(i \to j) = \min\{ 1, \exp(V_i - V_j) \}.
$
However,
in the first or global sampling process,
we choose the destination $j$
uniformly out of the $n$ bins.
%
In the second or local sampling process,
$j$ is either $i - 1$ or $i + 1$
(adjusted by the boundary condition)
with equal probability $1/2$.
%
Asymptotically,
the above global and local sampling processes
emulate the perfect and one-step processes
discussed in Appendix \ref{sec:Gamma},
respectively,
and we shall use
Eqs. \eqref{eq:Gamma_perfect}
and \eqref{eq:Gamma_onestep}
for the respective $\Gamma_k$'s
in the theoretical predictions.
%

@@

\begin{table}[h]
  \caption{\label{tab:error_Gaussian}
    Error of the Gaussian updating scheme of $\sigma = 10$.
    $t_0$ is given by Eq. \eqref{eq:t0_sinc}.
    \note{
      The data were collected in
      \texttt{data/tab1} by running
      the program \texttt{invt}.
    }%
  }
  \setlength{\tabcolsep}{5pt}
  \begin{tabular} { l | p{2cm} p{2.5cm} p{1.8cm} }
    \hline
      Schedule
    &
      Inverse-time, Eq. \eqref{eq:alpha_invt1}
    &
      Modified inverse-time,
      Eq.~\eqref{eq:alpha_invtlambda}
      with optimized $\lambda$
    &
      Optimal,
      Eq. \eqref{eq:q_opt}
    \\
    \hline
    Global
    &
    $4.4(1) \times 10^{-6}$
    &
    $7.1 \times 10^{-7}$
    &
    $2.6(1) \times 10^{-7}$
    \\
    \hline
    Local
    &
    $3.8(2) \times 10^{-4}$
    &
    $1.8(1) \times 10^{-4}$
    &
    $1.0(1) \times 10^{-4}$
    \\
    \hline
  \end{tabular}
\end{table}

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=1.0\linewidth]{fig/xerr.pdf}
  }
  \caption{
    \label{fig:xerr}
    Error components for the Gaussian updating scheme
    with $\sigma = 10$
    from the optimal schedule
    and the inverse-time schedule, Eq. \eqref{eq:alpha_invt1}.
    %
    The latter schedule failed to damp out the errors
    from the long-wavelength (local) eigenmodes
    with $K \ge 5$.
    %
    For longer simulations,
    the error components from the optimal schedule
    would roughly maintain the same shape,
    but join the initial error curve
    at a lower level (not shown).
    %
    \note{This figure was produced by \texttt{doc/fig/xerr.gp}.
      The data were saved under the directory \texttt{data/xerr}.
    }%
  }
\end{center}
\end{figure}

@@

To show Eq. \eqref{eq:varY},
we first use Eq. \eqref{eq:v_def}
to rewrite Eq. \eqref{eq:x_def}
as
$
  x_i
  =
  (V_i - \bar V) + \chi_i,
$
where
$
  \chi_i
  \equiv
  \ln \frac{ p_i } { p^*_i }
  -
  \sum_{ j = 1 }^n
    p_j \, \ln \frac{ p_j } { p^*_j }.
$
Thus, by Eqs. \eqref{eq:y_def} and \eqref{eq:Y_def},
we have
$$
y_k = Y_k + \sum_{i = 1}^n \phi_{k i} \, \chi_i.
$$
Since the second term on the right-hand side
is a constant, and the average of $y_k$ is zero,
we have
$$
\operatorname{var} Y_k
=
\operatorname{var} y_k
=
\langle y_k^2 \rangle.
$$
Finally, by using Eq. \eqref{eq:y2_eql},
we reach Eq. \eqref{eq:varY}.

@@

As an example, we used the method to compute
the integrals, $\Gamma_k$'s, from an MD simulation,
and compared them to those from
%Eq. \eqref{eq:Gamma_onestep}.
the local MC sampling process.
%
As shown in Fig. \ref{fig:gamcmp},
the larger values of $\Gamma_k$'s
with smaller $k$'s
from the two sampling processes
were almost proportional.
%
This suggests that
the optimal schedule
from simulations using local sampling processes
would be similar in general properties.

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/gamcmp.pdf}
  }
  \caption{
    \label{fig:gamcmp}
    Integrals of the autocorrelation functions
    from an MD simulation versus those
    from the local MC simulation.
    %
    Each simulation was ran for $10^{10}$ steps.
    %
    The bin size $\Delta z = 2\, \pi /n$ with $n = 100$.
    %
    For the MD simulation,
    the step size $\Delta t_\mathrm{MD} = 0.005$.
    %
    The velocity rescaling thermostat\cite{bussi2007} was
    applied for the unit reduced temperature
    every MD step with viscosity $20$.
    %
    The reflective boundary condition was used.
    %
    \note{The figure was produced by \texttt{doc/fig/gamcmp.gp}.
      The data were saved under \texttt{data/gamcmp}
      (see the \texttt{Makefile} there).
    }%
  }
\end{center}
\end{figure}
